---
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{authblk}
output: pdf_document
bibliography: pharmacogx.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\title{Biomarker discovery from large pharmacogenomics datasets}
\author[1,2]{Zhaleh Safikhani}
\author[1,2,3]{Benjamin Haibe-Kains}
\author[1, 2]{Petr Smrinov}
\author[1, 2]{Seyed Ali Madani Tonekaboni}
\affil[1]{Princess Margaret Cancer Centre, University Health Network, Toronto Canada}
\affil[2]{Department of Medical Biophysics, University of Toronto, Toronto Canada}
\affil[3]{Department of Computer Science, University of Toronto, Toronto Canada}

\maketitle
\tableofcontents

\section{Introduction}

Pharmacogenomics holds much potential to aid in discovering drug response
biomarkers and developing novel targeted therapies, leading to development of
precision medicine and working towards the goal of personalized therapy.
Several large experiments have been conducted, both to molecularly
characterize drug dose response across many cell lines, and to examine the
molecular response to drug administration. However, the experiments lack a
standardization of protocols and annotations, hindering meta-analysis across
several experiments.
*PharmacoGx* was developed to address these challenges, by providing a
unified framework for downloading and analyzing large pharmacogenomic datasets
which are extensively curated to ensure maximum overlap and consistency.
*PharmacoGx* is based on a level of abstraction from the raw
experimental data, and allows bioinformaticians and biologists to work with
data at the level of genes, drugs and cell lines. This provides a more
intuitive interface and, in combination with unified curation, simplifies
analyses between multiple datasets.

\subsection{Installation and Settings}
*PharmacoGx* requires that several packages are installed. However, all dependencies are available from CRAN or Bioconductor.
```{r install_pharmacogx, eval=FALSE}
source('http://bioconductor.org/biocLite.R')
biocLite('PharmacoGx')
```

Load *PharamacoGx* into your current workspace:
```{r loadlib, eval=TRUE, results='hide'}
library(PharmacoGx)
```
\subsection{Downloading PharmacoSet objects}
We have made the PharmacoSet objects of the curated datasets available for download using functions provided in the package. A table of available PharmacoSet objects can be obtained by using the *availablePSets* function. Any of the PharmacoSets in the table can then be downloaded by calling *downloadPSet*, which saves the datasets into a directory of the users choice, and returns the data into the R session. 
```{r download_psets, eval=TRUE, results='hide'}
  availablePSets()
  GDSC <- downloadPSet("GDSC") 
  CCLE <- downloadPSet("CCLE")
```

\section{Reproducibility}
*PharmacoGx* can be used to process pharmacogenomic datasets. First we want to check the heterogenity of cell lines in one of the available psets, CCLE.

```{r pie_chart, fig.cap="Tissue of origin of cell lines in CCLE study"}
mycol <- c("#8dd3c7","#ffffb3","#bebada","#fb8072","#80b1d3","#fdb462",
           "#b3de69","#fccde5","#d9d9d9","#bc80bd","#ccebc5","#ffed6f",
           "#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
           "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928")
pie(table(CCLE@cell[,"tissueid"]), 
    col=mycol, 
    main="Tissue types", 
    radius=1, 
    cex=0.8)
```

\subsection{Pharmacological profiles}
In pharmacogenomic studies Cells were also tested for their response to increasing concentrations of various compounds, 
and form this the IC50 and AUC were computed. These pharmacological profiles are available for all the psets in *PharmacoGx*.

```{r ccle_auc, fig.cap="Cells response to lapatinib in CCLE"}
CCLE.auc <- summarizeSensitivityProfiles(
                pSet=CCLE,
                sensitivity.measure="auc_published", 
                summary.stat="median",
                verbose=FALSE)
hist(CCLE.auc["lapatinib",], xlab="Cells response to lapatinib(AAC)", 
     col="gray", main="")
```

\subsection{Plotting Drug-Dose Response Data}
Drug-Dose response data included in the PharmacoSet objects can be conviniently plotted using the *drugDoseResponseCurve* function. Given a list of PharmacoSets, a drug name and a cell name, it will plot the drug dose response curves for the given cell-drug combination in each dataset, allowing direct comparisons of data between datasets. 

```{r curves, fig.cap="Cells response to lapatinib in CCLE"}
lapatinib.aac <- CCLE.auc["lapatinib",]
cells <- names(lapatinib.aac)[
                c(which.min(lapatinib.aac), 
                  which((lapatinib.aac > 0.2) & (lapatinib.aac < 0.4))[1],
                  which.max(lapatinib.aac))]
par(mfrow=c(2, 2))
drugDoseResponseCurve(drug="lapatinib", cellline=cells[1], 
                      pSets=CCLE, plot.type="Fitted", 
                      legends.label="auc_published")
drugDoseResponseCurve(drug="lapatinib", cellline=cells[2], 
                      pSets=CCLE, plot.type="Fitted", 
                      legends.label="auc_published")
drugDoseResponseCurve(drug="lapatinib", cellline=cells[3], 
                      pSets=CCLE, plot.type="Fitted", 
                      legends.label="auc_published")
```

\section{Replication}
In this section we will investigate the consistency between the GDSC and CCLE datasets. In both CCLE and GDSC, the transcriptome of cells was profiled using an Affymatrix microarray chip. Cells were also tested for their response to increasing concentrations of various compounds, and form this the IC50 and AUC were computed. However, the cell and drugs names used between the two datasets were not consistent. Furthermore, two different microarray platforms were used. However, *PharmacoGx* allows us to overcome these differences to do a comparative study between these two datasets. 

GDSC was profiled using the hgu133a platform, while CCLE was profiled with the expanded hgu133plus2 platform. While in this case the hgu133a is almost a strict subset of hgu133plus2 platform, the expression information in *PharmacoSet* objects is summarized by Ensemble Gene Ids, allowing datasets with different platforms to be directly compared. The probe to gene mapping is done using the BrainArray customCDF for each platform \cite{sabatti_thresholding_2002}.

To begin, you would load the datasets from disk or download them using the *downloadPSet* function above.

We want to investigate the consistency of the data between the two datasets. The common intersection between the datasets can then be found using *intersectPSet*. We create a summary of the gene expression and drug sensitivity measures for both datasets, so we are left with one gene expression profile and one sensitivity profile per cell line within each dataset. We can then compare the gene expression and sensitivity measures between the datasets using a standard correlation coefficient. 
```{r Replication_curves, fig.cap="Consistency of drug response curves across studies", results='hide'}
library(Biobase, verbose=FALSE)
common <- intersectPSet(pSets = list("CCLE"=CCLE, "GDSC"=GDSC), 
                        intersectOn = c("cell.lines", "drugs"), 
                        strictIntersect = TRUE)
drugs <- drugNames(common$CCLE)

##Example of concordant and discordant drug curves
cases <- rbind(
  c("CAL-85-1", "17-AAG"),
  c("HT-29", "PLX4720"),
  c("COLO-320-HSR", "AZD6244"),
  c("HT-1080", "PD-0332991"))

par(mfrow=c(2, 2))
for (i in 1:nrow(cases)) {
  drugDoseResponseCurve(pSets=common, 
                        drug=cases[i,2], 
                        cellline=cases[i,1], 
                        legends.label="ic50_published", 
                        plot.type="Fitted", 
                        ylim=c(0,130))
}
```

\subsection{Consistency of pharmacological profiles}
```{r sensitivity_scatter_plots, fig.height=5, fig.width=10, fig.cap="Concordance of AAC values"}
  
  ##AAC scatter plot 
GDSC.aac <- summarizeSensitivityProfiles(
    pSet=common$GDSC,
    sensitivity.measure='auc_recomputed', 
    summary.stat="median",
    verbose=FALSE)
CCLE.aac <- summarizeSensitivityProfiles(
  pSet=common$CCLE,
  sensitivity.measure='auc_recomputed', 
  summary.stat="median",
  verbose=FALSE)

GDSC.ic50 <- log10(summarizeSensitivityProfiles(
    pSet=common$GDSC, 
    sensitivity.measure='ic50_recomputed', 
    summary.stat="median",
    verbose=FALSE))
CCLE.ic50 <- log10(summarizeSensitivityProfiles(
  pSet=common$CCLE, 
  sensitivity.measure='ic50_recomputed', 
  summary.stat="median",
  verbose=FALSE))

  drug <- "lapatinib"
par(mfrow=c(1, 2))
plot(GDSC.aac[drug,], CCLE.aac[drug,], 
       xlab="GDSC", ylab="CCLE",
       main="lapatinib (AAC)", pch=20)
legend("topright",
         legend=sprintf("rs=%s", 
                 round(cor(GDSC.aac[drug,], 
                           CCLE.aac[drug,], 
                           method="spearman",
                           use="pairwise.complete.obs"), 
                       digits=2)), 
         bty="n")
plot(GDSC.ic50[drug,], CCLE.ic50[drug,], 
       xlab="GDSC", ylab="CCLE",
       main="lapatinib (IC50)", pch=20)
legend("topright",
         legend=sprintf("rs=%s", 
                 round(cor(GDSC.ic50[drug,], 
                           CCLE.ic50[drug,], 
                           method="spearman",
                           use="pairwise.complete.obs"), 
                       digits=2)), 
         bty="n")
```
\subsection{consistency assessment improved by Modified Concordance Index}
To better assess the concordance of multiple pharmacogenomic studies we introduced the modified concordance index (mCI). Recognizing that the noise in the drug screening assays is high and may yield to inaccurate sensitive-based ranking of cell lines with close AAC values, the mCI only considers cell line pairs with drug sensitivity (AAC) difference greater than $\delta$ .
```{r install_mCI, eval=FALSE}
library(devtools)
devtools::install_github("bhklab/mci", ref="light")
```

```{r mci, eval=TRUE, results='hide'}
library(mCI)
c_index <-  mc_index <- NULL
for(drug in drugs){
  tt <- mCI::paired.concordance.index(GDSC.aac[drug,], CCLE.aac[drug,], delta.pred=0, delta.obs=0, alternative="greater")
  c_index <- c(c_index, tt$cindex)
  tt <- mCI::paired.concordance.index(GDSC.aac[drug,], CCLE.aac[drug,], delta.pred=0.2, delta.obs=0.2, alternative="greater", logic.operator="or")
  mc_index <- c(mc_index, tt$cindex)
}
mp <- barplot(as.vector(rbind(c_index, mc_index)), beside=TRUE, col=c("blue", "red"), ylim=c(0, 1), ylab="concordance index", space=c(.15,.85), border=NA, main="mCI")
text(mp, par("usr")[3], labels=as.vector(rbind(drugs, rep("", 15))), srt=45, adj=c(1.1,1.1), xpd=TRUE, cex=.8)
abline(h=.7, lty=2)
```
\subsection{Known Biomarkers}
The association between molecular features and response to a given drug is modelled using a linear regression model adjusted for tissue source: 
$$Y = \beta_{0} + \beta_{i}G_i + \beta_{t}T + \beta_{b}B$$
where $Y$ denotes the drug sensitivity variable, $G_i$, $T$ and $B$ denote the expression of gene i, the tissue source and the experimental batch respectively, and $\beta$s are the regression coefficients. The strength of gene-drug association is quantified by $\beta_i$, above and beyond the relationship between drug sensitivity and tissue source. The variables Y and G are scaled (standard deviation equals to 1) to estimate standardized coefficients from the linear model. Significance of the gene-drug association is estimated by the statistical significance of $\beta_i$ (two-sided t test). P-values are then corrected for multiple testing using the false discovery rate (FDR) approach.

As an example of the reproducibility of biomarker discovery across pharmacogenomic studies, we can model the significance of the association between two drugs and their known biomarkers in CCLE and GDSC. We examine the association between drug *17-AAG* and gene *NQO1*, as well as drug *PD-0325901* and gene *BRAF*:



``` {r biomarker_discovery, results='asis'} 
  features <- PharmacoGx::fNames(CCLE, "rna")[
    which(featureInfo(CCLE,
                      "rna")$Symbol == "NQO1")]
    ccle.sig.rna <- drugSensitivitySig(pSet=CCLE, 
                                  mDataType="rna", 
                                  drugs=c("17-AAG"), 
                                  features=features, 
                                  sensitivity.measure="auc_published", 
                                  molecular.summary.stat="median", 
                                  sensitivity.summary.stat="median",
                                  verbose=FALSE)
    gdsc.sig.rna <- drugSensitivitySig(pSet=GDSC, 
                                  mDataType="rna", 
                                  drugs=c("17-AAG"), 
                                  features=features, 
                                  sensitivity.measure="auc_published", 
                                  molecular.summary.stat="median", 
                                  sensitivity.summary.stat="median",
                                  verbose=FALSE)
    ccle.sig.mut <- drugSensitivitySig(pSet=CCLE, 
                                  mDataType="mutation", 
                                  drugs=c("PD-0325901"), 
                                  features="BRAF", 
                                  sensitivity.measure="auc_published", 
                                  molecular.summary.stat="and", 
                                  sensitivity.summary.stat="median",
                                  verbose=FALSE)
    gdsc.sig.mut <- drugSensitivitySig(pSet=GDSC, 
                                  mDataType="mutation", 
                                  drugs=c("PD-0325901"), 
                                  features="BRAF", 
                                  sensitivity.measure="auc_published", 
                                  molecular.summary.stat="and", 
                                  sensitivity.summary.stat="median",
                                  verbose=FALSE)
    ccle.sig <- rbind(ccle.sig.rna, ccle.sig.mut)
    gdsc.sig <- rbind(gdsc.sig.rna, gdsc.sig.mut)
    known.biomarkers <- cbind("GDSC effect size"=gdsc.sig[,1], 
                              "GDSC pvalue"=gdsc.sig[,6], 
                              "CCLE effect size"=ccle.sig[,1], 
                              "CCLE pvalue"=ccle.sig[,6])
    rownames(known.biomarkers) <- c("17-AAG + NQO1","PD-0325901 + BRAF")
    library(xtable)
    xtable(known.biomarkers, digits=c(0, 2, -1, 2, -1), caption='Concordance of biomarkers across stuudies')
    par(mfrow=c(2, 2))
    CCLE_expr <- t(exprs(summarizeMolecularProfiles(CCLE, mDataType="rna", fill.missing=FALSE)))
    CCLE_cells <- intersect(rownames(CCLE_expr), colnames(CCLE.aac))
    plot(CCLE.aac["17-AAG", CCLE_cells], CCLE_expr[CCLE_cells, features],
     main="CCLE + 17-AAG + NQO1",
     cex.main=1, ylab="Predictions", xlab="drug sensitivity", pch=20, col="gray40")

    GDSC_expr <- t(exprs(summarizeMolecularProfiles(GDSC, mDataType="rna", fill.missing=FALSE)))
    GDSC_cells <- intersect(rownames(GDSC_expr), colnames(GDSC.aac))
    plot(GDSC.aac["17-AAG", GDSC_cells], GDSC_expr[GDSC_cells, features],
     main="GDSC + 17-AAG + NQO1",
     cex.main=1, ylab="Predictions", xlab="drug sensitivity", pch=20, col="gray40")
    
    CCLE_mut <- t(exprs(summarizeMolecularProfiles(CCLE, mDataType="mutation", fill.missing=FALSE, summary.stat="or")))
CCLE_cells <- intersect(rownames(CCLE_mut), colnames(CCLE.aac))
boxplot(CCLE.aac["PD-0325901", CCLE_cells]~ CCLE_mut[CCLE_cells, "BRAF"], col="gray80", pch=20, main="CCLE + PD-0325901 + BRAF",
        cex.main=1, xlab="mutation", ylab="drug sensitivity")

GDSC_mut <- t(exprs(summarizeMolecularProfiles(GDSC, mDataType="mutation", fill.missing=FALSE, summary.stat="or")))
GDSC_cells <- intersect(rownames(GDSC_mut), colnames(GDSC.aac))
boxplot(GDSC.aac["PD-0325901", GDSC_cells]~ GDSC_mut[GDSC_cells, "BRAF"], col="gray80", pch=20, main="GDSC + PD-0325901 + BRAF",
        cex.main=1, xlab="mutation", ylab="drug sensitivity")

```
Sactter Plot for know biomarkers (4 plots)

\subsection{Machne Learning and Biomarker Discovery}
Some of the widely used multivariate machine learning methods such as elastic net, Random Forest (RF) and Support Vector Machine (SVM) have been already implemented in the MLWorkshop. It optimizes hyperparameters of these methods in the training phase. To assess the performance of the predictive models, it implements *m* number of sampling with *n-fold* cross validations (CV). The performance will then be assessed by multiple metrics including pearson correlation coefficient, concordance index and modified concordance index. 

```{r install_PharmacoGxML and dependencies, eval=FALSE}
library(devtools)
install_github("bhklab/PharmacoGx-ML")
```

```{r machine_learning, results='asis'} 
library(mRMRe, verbose=FALSE)
library(Biobase, verbose=FALSE)
library(PharmacoGxML, verbose=FALSE)
library(Hmisc, verbose=FALSE)
library(glmnet, verbose=FALSE)
library(caret, verbose=FALSE)

##Preparing trainig dataset
train_expr <- t(exprs(summarizeMolecularProfiles(GDSC, mDataType="rna", fill.missing=FALSE, verbose=FALSE)))
aac <- summarizeSensitivityProfiles(GDSC, sensitivity.measure="auc_recomputed", drug="lapatinib", fill.missing=FALSE, verbose=FALSE)
cells <- intersect(rownames(train_expr), names(aac))
df <- as.matrix(cbind(train_expr[cells,], "lapatinib"=aac[cells]))


##Preparing validation dataset
validation_expr <- summarizeMolecularProfiles(CCLE, mDataType="rna", fill.missing=FALSE, verbose=FALSE)
actual_labels <- summarizeSensitivityProfiles(CCLE, sensitivity.measure="auc_recomputed", drug="lapatinib", fill.missing=FALSE, verbose=FALSE)


par(mfrow=c(4, 2))
for(method in c("ridge", "lasso", "random_forest", "svm")){
  res <- optimization(train=df[, -ncol(df), drop=F],
                      labels=t(df[, ncol(df), drop=F]),
                      method=method,
                      folds.no=5,
                      sampling.no=1,
                      features.no=100,
                      feature.selection="mRMR",
                      assessment=c("corr", "mCI"))
  
  validation_labels <- prediction(model=res$model$lapatinib,
                                  validation.set=t(exprs(validation_expr)),
                                  validation.labels=actual_labels,
                                  method=method,
                                  assessment="mCI")

}

```
\section{Limitations/ Future direction}

\section{Session Info}
This document was generated with the following R version and packages loaded:
```{r sessionInfo,echo=FALSE, results='asis'}
  toLatex(sessionInfo())
```



